---
title: "Exploring *Bacteroides*-Proteobacteria association"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

### Relative abundance dynamics of *Bacteroides* and pathogenic Proteobacteria

We are going to visualize the change in relative abundance of *Bacteroides* and pathogenic Proteobacteria over time in the allo-HCT patient dataset. We want to determine if the dynamics are different in patients that suffered infections by the Proteobacteria compared to patients who did not get infected. For pathogens, we already expect their abundance increases and we are mostly replicating some of the results of Liao et al, 2021.

```{r}
# Load required libraries and functions
suppressMessages(library(tidyverse))
source("code/rfunctions/data_helpers.r")

# Prepare genus-level counts and metadata for plotting
tblrel_genus_meta <- prep_taxa_counts("genus")

# Plot the average relative abundance of Escherichia over time, grouped by infection status
tblrel_genus_meta %>%
  filter(!is.na(DayRelativeToNearestHCT), DayRelativeToNearestHCT >= -10, DayRelativeToNearestHCT <= 30) %>%
  group_by(DayRelativeToNearestHCT, Escherichia_infection) %>%
  summarise(avg_abundance = mean(`Escherichia-Shigella_abund`, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = DayRelativeToNearestHCT, y = avg_abundance, color = Escherichia_infection)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(`TRUE` = "#EE2C2C", `FALSE` = "gray60")) +
  labs(
  title = "Average relative abundance of escherichia over time",
   x = "Days Relative to Nearest HCT",
   y = "Average Relative Abundance of Escherichia",
   color = "Escherichia infection"
  ) +
  custom_theme +
  scale_y_continuous(limits = c(0, 1))    
    
# Plot the average relative abundance of Bacteroides over time, grouped by infection status
tblrel_genus_meta %>%
  filter(!is.na(DayRelativeToNearestHCT), DayRelativeToNearestHCT >= -15, DayRelativeToNearestHCT <= 30) %>%
  group_by(DayRelativeToNearestHCT, Escherichia_infection) %>%
  summarise(avg_abundance = mean(Bacteroides_abund, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = DayRelativeToNearestHCT, y = avg_abundance, color = Escherichia_infection)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(`TRUE` = "#EE2C2C", `FALSE` = "gray60")) +
  labs(
   title = "Average Relative Abundance of Bacteroides",
   x = "Days Relative to Nearest HCT",
   y = "Average relative abundance of Bacteroides over time",
   color = "Escherichia infection"
      ) +
  custom_theme +
  scale_y_continuous(limits = c(0, 1))

# Plot the average relative abundance of Klebsiella over time, grouped by infection status
tblrel_genus_meta %>%
  filter(!is.na(DayRelativeToNearestHCT), DayRelativeToNearestHCT >= -15, DayRelativeToNearestHCT <= 30) %>%
   group_by(DayRelativeToNearestHCT, Klebsiella_infection) %>%
   summarise(avg_abundance = mean(Klebsiella_abund, na.rm = TRUE), .groups = "drop") %>%
   ggplot(aes(x = DayRelativeToNearestHCT, y = avg_abundance, color = Klebsiella_infection)) +
   geom_line(linewidth = 1) +
   scale_color_manual(values = c(`TRUE` = "#EE2C2C", `FALSE` = "gray60")) +
   labs(
    title = "Average Relative Abundance of Klebsiella Over Time",
    x = "Days Relative to Nearest HCT",
    y = "Average relative abundance of Klebsiella over time",
      color = "Klebsiella infection"
      ) +
  custom_theme +
  scale_y_continuous(limits = c(0, 1))

# Plot the average relative abundance of Bacteroides over time, grouped by infection status
tblrel_genus_meta %>%
  filter(!is.na(DayRelativeToNearestHCT), DayRelativeToNearestHCT >= -15, DayRelativeToNearestHCT <= 30) %>%
  group_by(DayRelativeToNearestHCT, Klebsiella_infection) %>%
  summarise(avg_abundance = mean(Bacteroides_abund, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = DayRelativeToNearestHCT, y = avg_abundance, color = Klebsiella_infection)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(`TRUE` = "#EE2C2C", `FALSE` = "gray60")) +
  labs(
   title = "Average Relative Abundance of Bacteroides Over Time",
   x = "Days Relative to Nearest HCT",
   y = "Average relative abundance of Bacteroides over time",
    color = "Klebsiella infection"
      ) +
  custom_theme +
  scale_y_continuous(limits = c(0, 1))

# Plot the average relative abundance of Streptococcus over time, grouped by infection status
tblrel_genus_meta %>%
  filter(!is.na(DayRelativeToNearestHCT), DayRelativeToNearestHCT >= -15, DayRelativeToNearestHCT <= 30) %>%
   group_by(DayRelativeToNearestHCT, Klebsiella_infection) %>%
   summarise(avg_abundance = mean(Streptococcus_abund, na.rm = TRUE), .groups = "drop") %>%
   ggplot(aes(x = DayRelativeToNearestHCT, y = avg_abundance, color = Klebsiella_infection)) +
   geom_line(linewidth = 1) +
   scale_color_manual(values = c(`TRUE` = "#EE2C2C", `FALSE` = "gray60")) +
   labs(
    title = "Average Relative Abundance of Streptococcus Over Time",
    x = "Days Relative to Nearest HCT",
    y = "Average relative abundance of Streptococcus over time",
      color = "Klebsiella infection"
      ) +
  custom_theme +
  scale_y_continuous(limits = c(0, 1))
```

There is no clear evidence that the relative abundance of *Bacteroides* is following a similar dynamic to that of *Escherichia*, or even that there is a higher relative abundance of Baceroides in patients who became infected with Escherichia. 

Interestingly, there is not an increase in the relative abundance of *Klebsiella* in patients that eventually developed a *Klebsiella* infection. This is a bit counter-intuitive and I need to examine the time series for those patients more closely.

### Absolute abundance dynamics of *Bacteroides* and pathogenic Proteobacteria

It is possible that the link between the dynamics of *Bacteroides* and pathogenic proteobacteria is not evident when looking at relative abundance due to non-independence. So, we are going to now look at the dynamics in the subset of the samples for which we have qPCR data for 16S, which allows us to convert to absolute abundances.

```{r}
# Load the qpcr data
tblqpcr <- suppressMessages(read_csv("data/tblqpcr.csv"))

# Get table with genus absolute abundance
tblabs_genus_meta <- tblrel_genus_meta %>%
  filter(Sample %in% unique(tblqpcr$SampleID)) %>%
  left_join(tblqpcr, by = c("Sample" = "SampleID")) %>%
  mutate(across(ends_with("_abund"), ~ .x * qPCR16S))

# Plot the log of absolute abundance of Escherichia over time, grouped by infection status
tblabs_genus_meta %>%
  filter(!is.na(DayRelativeToNearestHCT), DayRelativeToNearestHCT >= -10, DayRelativeToNearestHCT <= 30) %>%
  group_by(DayRelativeToNearestHCT, Escherichia_infection) %>%
  summarise(avg_abs_abundance = mean(`Escherichia-Shigella_abund`, na.rm = TRUE), .groups = "drop") %>%
  mutate(log_avg_abs_abundance = log10(avg_abs_abundance)) %>%
  ggplot(aes(x = DayRelativeToNearestHCT, y = log_avg_abs_abundance, color = Escherichia_infection)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(`TRUE` = "#EE2C2C", `FALSE` = "gray60")) +
  labs(
    title = "Log of Average Absolute Abundance of Escherichia Over Time",
    x = "Days Relative to Nearest HCT",
    y = "Log Average Absolute Abundance of Escherichia",
    color = "Escherichia infection"
    ) +
  custom_theme

# Plot the log of absolute abundance of Bacteroides over time, grouped by Escherichia infection status
tblabs_genus_meta %>%
  filter(!is.na(DayRelativeToNearestHCT), DayRelativeToNearestHCT >= -10, DayRelativeToNearestHCT <= 30) %>%
  group_by(DayRelativeToNearestHCT, Escherichia_infection) %>%
  summarise(avg_abs_abundance = mean(Bacteroides_abund, na.rm = TRUE), .groups = "drop") %>%
  mutate(log_avg_abs_abundance = log10(avg_abs_abundance)) %>%
  ggplot(aes(x = DayRelativeToNearestHCT, y = log_avg_abs_abundance, color = Escherichia_infection)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(`TRUE` = "#EE2C2C", `FALSE` = "gray60")) +
  labs(
    title = "Log of Average Absolute Abundance of Bacteroides Over Time",
    x = "Days Relative to Nearest HCT",
    y = "Log Average Absolute Abundance of Bacteroides",
    color = "Escherichia infection"
    ) +
  custom_theme
```

Absolute abundance is not as useful to examine the dynamics because there is a lot of missing data for infected patients. This is why the red lines drop to 0 in the plots. In any case, there is no indication that the dynamic pattern is different from the relative abundance based on this data.

### Patient-specific time series


```{r, echo = FALSE, eval = FALSE}
# Prepare asv-level counts and metadata for plotting
tblrel_asv_meta <- prep_taxa_counts("asv")

# Load taxonomy data with colors and filter to tax level
taxlevel_color <- read_csv("data/tblASVtaxonomy_silva132_v4v5_filter.csv") %>%
  select(ASV, HexColor, ColorOrder) %>%
  distinct(ASV, .keep_all = TRUE)

# Filter data for PatientID 1511 and time points between -5 and 20
patient_1511_data <- tblrel_asv_meta %>%
  filter(PatientID == 1511, DayRelativeToNearestHCT >= -5, DayRelativeToNearestHCT <= 20) %>%
  select(DayRelativeToNearestHCT, ends_with("_abund")) %>%
  pivot_longer(
    cols = ends_with("_abund"),
    names_to = "Taxa",
    values_to = "RelativeAbundance"
  ) %>%
  mutate(Taxa = str_remove(Taxa, "_abund"))

# Merge patient data with taxonomy colors and order by ColorOrder
patient_1511_data <- patient_1511_data %>%
  left_join(taxlevel_color, by = c("Taxa" = "ASV")) %>%
  arrange(ColorOrder)

# Plot microbial composition for PatientID 1511 with assigned colors and order
ggplot(patient_1511_data, aes(x = DayRelativeToNearestHCT, y = RelativeAbundance, fill = reorder(Taxa, -ColorOrder))) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = setNames(patient_1511_data$HexColor, patient_1511_data$Taxa)) +
  labs(
    title = "Microbial Composition of Patient 1511",
    x = "Days Relative to Nearest HCT",
    y = "Relative Abundance"
  ) +
  custom_theme +
  theme(legend.position = "none")
```

### Cooccurrence

```{r}
# Transform the table into a binary presence-absence matrix
binary_matrix_full <- tblrel_genus_meta %>%
  select(ends_with("_abund")) %>%
  mutate(across(everything(), ~ ifelse(. >= 0.001, 1, 0))) %>%
  as.matrix()
binary_matrix_infected <- tblrel_genus_meta %>%
  filter(Escherichia_infection == TRUE) %>%
  select(ends_with("_abund")) %>%
  mutate(across(everything(), ~ ifelse(. >= 0.001, 1, 0))) %>%
  as.matrix()

# Calculate the cooccurrence matrix
cooccurrence_matrix_full <- t(binary_matrix_full) %*% binary_matrix_full %>%
  as.matrix()
cooccurrence_matrix_infected <- t(binary_matrix_infected) %*% binary_matrix_infected %>%
  as.matrix()

# Frequency of Bacteroides and Escherichia in the samples
bacteroides_freq_full <- cooccurrence_matrix_full["Bacteroides_abund", "Bacteroides_abund"]
escherichia_freq_full <- cooccurrence_matrix_full["Escherichia-Shigella_abund", "Escherichia-Shigella_abund"]
bacteroides_freq_infected <- cooccurrence_matrix_infected["Bacteroides_abund", "Bacteroides_abund"]
escherichia_freq_infected <- cooccurrence_matrix_infected["Escherichia-Shigella_abund", "Escherichia-Shigella_abund"]

# Frequencies of Bacteroides and Escherichia cooccurrence
cooccurrence_freq_full <- cooccurrence_matrix_full["Bacteroides_abund", "Escherichia-Shigella_abund"]
cooccurrence_freq_infected <- cooccurrence_matrix_infected["Bacteroides_abund", "Escherichia-Shigella_abund"]
```

 - *Escherichia* occurs in `r escherichia_freq_full` or `r (escherichia_freq_full/nrow(binary_matrix_full))*100`% of the samples, and in `r escherichia_freq_infected` or `r (escherichia_freq_infected/nrow(binary_matrix_infected))*100`% of the samples from patients who had positive blood cultures of *E. coli*
 - *Bacteroides* occurs in `r bacteroides_freq_full` or `r (bacteroides_freq_full/nrow(binary_matrix_full))*100`% of the samples, and in `r bacteroides_freq_infected` or `r (bacteroides_freq_infected/nrow(binary_matrix_infected))*100`% of the samples from patients who had positive blood cultures of *E. coli*
 - Bacteroides and *Escherichia* cooccur in `r cooccurrence_freq_full` samples, which corresponds to `r (cooccurrence_freq_full/escherichia_freq_full)*100`% of the samples with *Escherichia* and `r (cooccurrence_freq_full/bacteroides_freq_full)*100`% of the samples with *Bacteroides*.
 - Bacteroides and *Escherichia* cooccur in `r cooccurrence_freq_infected` samples from infected patients, which corresponds to `r (cooccurrence_freq_infected/escherichia_freq_infected)*100`% of the samples from infected patients colonized with *Escherichia* and `r (cooccurrence_freq_infected/bacteroides_freq_infected)*100`% of the samples infected patients colonized with  with *Bacteroides*.