---
title: "Exploring clinical and microbiome data from allo-HCT patients (Liao et al 2021)"
format: html
---

This is my initial exploration of the data from Liao et al 2021.

```{r}
# Load libraries
library(tidyverse, quietly = TRUE)
```

## Clinical data

### Transplant metadata

We first explored the transplant metadata:

```{r}
tblhctmeta <- read_csv("data/tblhctmeta.csv")
head(tblhctmeta)
```

```{r, echo = FALSE}
# EXPLORING TRANSPLANT METADATA

# Total number of patients
n_patients <- tblhctmeta$PatientID %>% unique() %>% length()

# Patients with 1 HCTS
n_patients_1hct <- tblhctmeta %>% 
    group_by(PatientID) %>% 
    filter(n() == 1) %>% 
    pull(PatientID) %>% 
    unique() %>%
    length()

# Patients with 2 HCTs
n_patients_2hct <- tblhctmeta %>% 
    group_by(PatientID) %>% 
    filter(n() == 2) %>% 
    pull(PatientID) %>% 
    unique() %>%
    length()

# Patients with 3 HCTs
n_patients_3hct <- tblhctmeta %>% 
    group_by(PatientID) %>% 
    filter(n() == 3) %>% 
    pull(PatientID) %>% 
    unique() %>%
    length()

# Calculate the time differences for patients with more than 1 HCT
time_diffs <- tblhctmeta %>%
    group_by(PatientID) %>%
    filter(n() > 1) %>%
    arrange(PatientID, TimepointOfTransplant) %>%
    mutate(time_diff = TimepointOfTransplant - lag(TimepointOfTransplant)) %>%
    filter(!is.na(time_diff)) %>%
    pull(time_diff)

# Minimum, median, and maximum time differences
min_time_diff <- min(time_diffs)
median_time_diff <- median(time_diffs)
max_time_diff <- max(time_diffs)

# EXPLORING DISEASE DATA

# Filter the dataset to only consider one record for each patient
unique_patients <- tblhctmeta %>%
    group_by(PatientID) %>%
    slice(1) %>%
    ungroup()

# Define a custom theme for plots
custom_theme <- theme(
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.75), 
    axis.text = element_text(color = "black", size = 10)
)

# Create a bar plot showing the frequency of each disease
disease_plot <- unique_patients %>%
    filter(!is.na(Disease)) %>%
    count(Disease) %>%
    ggplot(aes(x = reorder(Disease, -n), y = n)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = n), vjust = -0.5) +
    ylim(0, 700) +
    labs(x = "Disease",
         y = "Frequency") +
    custom_theme +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

# EXPLORING TRANSPLANT SOURCE

# Create a bar plot showing the frequency of each HCT source
hct_source_plot <- unique_patients %>%
    filter(!is.na(HCTSource)) %>%
    count(HCTSource) %>%
    ggplot(aes(x = reorder(HCTSource, -n), y = n)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = n), vjust = -0.5) +
    ylim(0, 550) +
    labs(x = "HCT Source",
         y = "Frequency") +
    scale_x_discrete(labels = c("TCD" = "T-Cell depleted", 
                                "PBSC_unmodified" = "Peripheral blood stem cells", 
                                "cord" = "Cord blood", 
                                "BM_unmodified" = "Bone marrow")) +
    custom_theme +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

# EXPLORING CROSS-STUDY LINKS

# How many patients of these patients were included in previous studies?
n_patients_na_wbc <- tblhctmeta %>%
    filter(!is.na(wbcPatientId)) %>%
    pull(PatientID) %>%
    unique() %>%
    length()
n_patients_na_fmt <- tblhctmeta %>%
    filter(is.na(autoFmtPatientId)) %>%
    pull(PatientID) %>%
    unique() %>%
    length()
n_patients_na_nejm <- tblhctmeta$nejmPatientId %>% sum()
```

This table contains data for HCTs in `r n_patients` patients. Of those, `r n_patients_1hct` received 1 HCT, `r n_patients_2hct` received 2 HCTs, and `r n_patients_3hct` received 3 HCTs. The minimum time between two HCTs for the same patient was `r min_time_diff` days, the maximum was `r max_time_diff` days, and the median was `r median_time_diff` days.

The most commonm disease among these patients was Leukemia, with `r unique_patients %>% filter(Disease == "Leukemia") %>% pull() %>% length()` cases. Here's the frequency distribution of all diseases: 

```{r, echo = FALSE}
print(disease_plot)
```

The most common source for the HCT was T-Cell depleted, used in `r unique_patients %>% filter(HCTSource == "TCD") %>% pull() %>% length()` transplants. Here's the freqeuncy distribution of the transplant sources:

```{r, echo = FALSE}
print(hct_source_plot)
```

Finally, all `r n_patients_na_wbc` patients were included in a [previous study on the role of the microbiota in  white blood cell dynamics](https://doi.org/10.1038/s41586-020-2971-8); `r n_patients_na_fmt` of the patients were included in a [previous study describing an autologous faecal microbiota trial](https://doi.org/10.1126/scitranslmed.aap9489); and `r n_patients_na_nejm` of the patients were part of a [previous study linking microbiome diversity to mortality during HCT](https://www.nejm.org/doi/full/10.1056/NEJMoa1900623).
 
***Issues to check***

- Some patients have NA in the Disease column for one of the HCT records even when another record specified a disease–-why?

- What is the actual source for the TCD cases?