---
title: "Exploring clinical and microbiome data from allo-HCT patients (Liao et al 2021)"
format: html
---

This is my initial exploration of the data from Liao et al 2021.

```{r}
# Load libraries
library(tidyverse, quietly = TRUE)
```

## Clinical data

### Transplant metadata

We first explored the transplant metadata:

```{r}
tblhctmeta <- read_csv("data/tblhctmeta.csv")
head(tblhctmeta)
```

```{r, echo = FALSE}
# EXPLORING TRANSPLANT METADATA

# Total number of patients
n_patients <- tblhctmeta$PatientID %>% unique() %>% length()

# Patients with 1 HCTS
n_patients_1hct <- tblhctmeta %>% 
    group_by(PatientID) %>% 
    filter(n() == 1) %>% 
    pull(PatientID) %>% 
    unique() %>%
    length()

# Patients with 2 HCTs
n_patients_2hct <- tblhctmeta %>% 
    group_by(PatientID) %>% 
    filter(n() == 2) %>% 
    pull(PatientID) %>% 
    unique() %>%
    length()

# Patients with 3 HCTs
n_patients_3hct <- tblhctmeta %>% 
    group_by(PatientID) %>% 
    filter(n() == 3) %>% 
    pull(PatientID) %>% 
    unique() %>%
    length()

# Calculate the time differences for patients with more than 1 HCT
time_diffs <- tblhctmeta %>%
    group_by(PatientID) %>%
    filter(n() > 1) %>%
    arrange(PatientID, TimepointOfTransplant) %>%
    mutate(time_diff = TimepointOfTransplant - lag(TimepointOfTransplant)) %>%
    filter(!is.na(time_diff)) %>%
    pull(time_diff)

# Minimum, median, and maximum time differences
min_time_diff <- min(time_diffs)
median_time_diff <- median(time_diffs)
max_time_diff <- max(time_diffs)

# EXPLORING DISEASE DATA

# Filter the dataset to only consider one record for each patient
unique_patients <- tblhctmeta %>%
    group_by(PatientID) %>%
    slice(1) %>%
    ungroup()

# Define a custom theme for plots
custom_theme <- theme(
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.75), 
    axis.text = element_text(color = "black", size = 10)
)

# Create a bar plot showing the frequency of each disease
disease_plot <- unique_patients %>%
    filter(!is.na(Disease)) %>%
    count(Disease) %>%
    ggplot(aes(x = reorder(Disease, -n), y = n)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = n), vjust = -0.5) +
    ylim(0, 700) +
    labs(x = "Disease",
         y = "Frequency") +
    custom_theme +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

# EXPLORING TRANSPLANT SOURCE

# Create a bar plot showing the frequency of each HCT source
hct_source_plot <- unique_patients %>%
    filter(!is.na(HCTSource)) %>%
    count(HCTSource) %>%
    ggplot(aes(x = reorder(HCTSource, -n), y = n)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = n), vjust = -0.5) +
    ylim(0, 550) +
    labs(x = "HCT Source",
         y = "Frequency") +
    scale_x_discrete(labels = c("TCD" = "T-Cell depleted", 
                                "PBSC_unmodified" = "Peripheral blood stem cells", 
                                "cord" = "Cord blood", 
                                "BM_unmodified" = "Bone marrow")) +
    custom_theme +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

# EXPLORING CROSS-STUDY LINKS

# How many patients of these patients were included in previous studies?
n_patients_na_wbc <- tblhctmeta %>%
    filter(!is.na(wbcPatientId)) %>%
    pull(PatientID) %>%
    unique() %>%
    length()
n_patients_na_fmt <- tblhctmeta %>%
    filter(is.na(autoFmtPatientId)) %>%
    pull(PatientID) %>%
    unique() %>%
    length()
n_patients_na_nejm <- tblhctmeta$nejmPatientId %>% sum()
```

This table contains data for HCTs in `r n_patients` patients. Of those, `r n_patients_1hct` received 1 HCT, `r n_patients_2hct` received 2 HCTs, and `r n_patients_3hct` received 3 HCTs. The minimum time between two HCTs for the same patient was `r min_time_diff` days, the maximum was `r max_time_diff` days, and the median was `r median_time_diff` days.

The most commonm disease among these patients was Leukemia, with `r unique_patients %>% filter(Disease == "Leukemia") %>% pull() %>% length()` cases. Here's the frequency distribution of all diseases: 

```{r, echo = FALSE}
print(disease_plot)
```

The most common source for the HCT was T-Cell depleted, used in `r unique_patients %>% filter(HCTSource == "TCD") %>% pull() %>% length()` transplants. Here's the freqeuncy distribution of the transplant sources:

```{r, echo = FALSE}
print(hct_source_plot)
```

Finally, all `r n_patients_na_wbc` patients were included in a [previous study on the role of the microbiota in  white blood cell dynamics](https://doi.org/10.1038/s41586-020-2971-8); `r n_patients_na_fmt` of the patients were included in a [previous study describing an autologous faecal microbiota trial](https://doi.org/10.1126/scitranslmed.aap9489); and `r n_patients_na_nejm` of the patients were part of a [previous study linking microbiome diversity to mortality during HCT](https://www.nejm.org/doi/full/10.1056/NEJMoa1900623).
 
***Issues to check***

- Some patients have NA in the Disease column for one of the HCT records even when another record specified a disease–-why?

- What is the actual source for the TCD cases?

### Anti-infective drug metadata

Then we looked at the data on anti-infective drug administration.

```{r}
tbldrug <- read_csv("data/tbldrug.csv")
head(tbldrug)
```

We made sure all patients in the antibiotic data are present in the transplant metadata:

```{r, echo = FALSE}
# Check if all PatientIDs in tbldrug are present in tblhctmeta
missing_patient_ids <- tbldrug %>%
    filter(!PatientID %in% tblhctmeta$PatientID) %>%
    pull(PatientID) %>%
    unique()

# Output the result
if (length(missing_patient_ids) == 0) {
    message("All PatientIDs in tbldrug are present in tblhctmeta.")
} else {
    message("The following PatientIDs in tbldrug are not present in tblhctmeta: ", paste(missing_patient_ids, collapse = ", "))
}
```

The table includes data for `r tbldrug %>% pull(Category) %>% unique() %>% length()` categories of anti-infectives:

```{r, echo = FALSE}
tbldrug %>% pull(Category) %>% unique()
```

Here's a table with anti-infective drugs administered to the patients:

```{r, echo = FALSE}
# Table of unique drugs, their categories, and the number of patients who received them
unique_drugs <- tbldrug %>%
    group_by(Factor, Category) %>%
    summarise(n_patients = n_distinct(PatientID)) %>%
    arrange(desc(n_patients), Category, Factor)

print(unique_drugs, n = 82)
```

Next, we were interested in the timeline of administration of anti-infective drugs. Specifically, we wanted to know if there were time periods relative to the transplant where most patients were under a similar anti-infective drug regime. This could guide the selection of specific time frames where model predicitons of blood stream infections might be more accurate and clinically relevant.

```{r, echo = FALSE}
# Load RColorBrewer for color palettes
library(RColorBrewer)

# Filter patients with a single HCT
single_hct_patients <- tblhctmeta %>%
        group_by(PatientID) %>%
        filter(n() == 1) %>%
        pull(PatientID)

# Most common drugs for plotting
top_drugs <- tbldrug %>%
    group_by(Category) %>%
    filter(n() > 400) %>%
    pull(Category) %>%
    unique()

# Filter the drug data for these patients
single_hct_drug_data <- tbldrug %>%
        filter(PatientID %in% single_hct_patients & Category %in% top_drugs)

# Filter the data to the specified range
filtered_drug_data_oral <- single_hct_drug_data %>%
        filter(StartDayRelativeToNearestHCT > -30 & StopDayRelativeToNearestHCT < 30 & Route == "oral")

# Filter the data to the specified range
filtered_drug_data_iv <- single_hct_drug_data %>%
        filter(StartDayRelativeToNearestHCT > -30 & StopDayRelativeToNearestHCT < 30 & Route == "intravenous")

# Define a color palette with 21 different colors
color_palette <- brewer.pal(n = 11, name = "Paired")
#if (length(unique(single_hct_drug_data$Category)) > 9) {
#    color_palette <- colorRampPalette(color_palette)(length(unique(single_hct_drug_data$Category)))
#}

# Create a dumbbell plot with each PatientID/Category combination as a different row
drug_timeline_iv <- filtered_drug_data_iv %>%
    ggplot(aes(y = interaction(PatientID, Category), x = StartDayRelativeToNearestHCT, xend = StopDayRelativeToNearestHCT, color = Category)) +
    geom_segment(aes(x = StartDayRelativeToNearestHCT, xend = StopDayRelativeToNearestHCT, y = interaction(PatientID, Category), yend = interaction(PatientID, Category)), linewidth = 0.08) +
    scale_color_manual(values = setNames(color_palette, sort(unique(filtered_drug_data_iv$Category))), breaks = sort(unique(filtered_drug_data_iv$Category), decreasing = TRUE)) +
    labs(title = "Intravenous anti-infective drugs",
        x = "Time Relative to Nearest HCT (days)",
         y = "Patient ID / Drug Category",
         color = "Drug Category") +
    scale_x_continuous(breaks = c(-30, -20, -10, -2, 0, 7, 10, 12, 20, 30)) +
    custom_theme +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.background = element_rect(fill = "transparent"),
          legend.key = element_rect(fill = "transparent"),
          legend.key.size = unit(1.5, "lines")) +
    guides(color = guide_legend(override.aes = list(linewidth = 3))) +
    geom_vline(xintercept = c(-10, -2, 7, 12), linetype = "dashed")

# Create a dumbbell plot with each PatientID/Category combination as a different row
drug_timeline_oral <- filtered_drug_data_oral %>%
    ggplot(aes(y = interaction(PatientID, Category), x = StartDayRelativeToNearestHCT, xend = StopDayRelativeToNearestHCT, color = Category)) +
    geom_segment(aes(x = StartDayRelativeToNearestHCT, xend = StopDayRelativeToNearestHCT, y = interaction(PatientID, Category), yend = interaction(PatientID, Category)), linewidth = 0.08) +
    scale_color_manual(values = setNames(color_palette, sort(unique(filtered_drug_data_iv$Category))), breaks = sort(unique(filtered_drug_data_iv$Category), decreasing = TRUE)) +
    labs(title = "Oral anti-infective drugs",
        x = "Time Relative to Nearest HCT (days)",
         y = "Patient ID / Drug Category",
         color = "Drug Category") +
    scale_x_continuous(breaks = c(-30, -20, -10, -2, 0, 7, 10, 12, 20, 30)) +
    custom_theme +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.background = element_rect(fill = "transparent"),
          legend.key = element_rect(fill = "transparent"),
          legend.key.size = unit(1.5, "lines")) +
    guides(color = guide_legend(override.aes = list(linewidth = 3))) +
    geom_vline(xintercept = c(-10, -2, 7, 12), linetype = "dashed")

#ggsave("plot.pdf")
```

There does seem to be some time frames where the anti-infective drugs administered are highly similar across patients:

```{r, echo = FALSE}
print(drug_timeline_iv)
print(drug_timeline_oral)
```