---
title: "Ecology of *E. faecium* dominations and BSI risk"
engine: knitr
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

## Role of antibiotics on BSI risk

Let's start by preparing the clinical and microbiome datasets:

```{r}
# Load required libraries and functions
suppressMessages(library(tidyverse))
source("code/rfunctions/data_helpers.r")

# Prepare genus-level counts
tblrel_genus_meta <- prep_taxa_counts("genus")

# Load table with qPCR data
tblqpcr <- read_csv("data/tblqpcr.csv")

# Summarize patients with Enterococcus dominations and infections
domination_bsi_summary <- tblrel_genus_meta %>%
  group_by(PatientID) %>%
  summarize(
    ever_dominated = any(Enterococcus_abund >= 0.10, na.rm = TRUE),
    ever_infected = any(Enterococcus_infection == TRUE, na.rm = TRUE)
  )

# List of dominated patients
dominated_patients <- domination_bsi_summary %>% 
    filter(ever_dominated == TRUE) %>%
    pull(PatientID)

# Load patient-day table with clinical covariates and filter to days -9 to 30
load("analyses/processed_data/tblpatientday_clinical.RData")
tblpatientday_clinical <- tblpatientday_clinical %>%
    drop_na() %>%
    filter(between(as.numeric(as.character(DayRelativeToNearestHCT)), -9, 30)) %>%
    arrange(PatientID, DayRelativeToNearestHCT) %>%
    group_by(PatientID) %>%
    mutate(Enterococcus_faecium_infection = if_else(
      Enterococcus_faecium_infection == 1 & cumsum(Enterococcus_faecium_infection) == 1, 1, 0
  )) %>%
  ungroup()

# Make a version of the patient day table that includes the microbiome data
tblpatientday_microbiome <- tblpatientday_clinical %>%
    mutate(DayRelativeToNearestHCT = as.numeric(as.character(DayRelativeToNearestHCT))) %>%
    left_join(tblrel_genus_meta, by = c("PatientID", "DayRelativeToNearestHCT")) %>%
    mutate(DayRelativeToNearestHCT = as.factor(DayRelativeToNearestHCT)) %>%
    drop_na()

# Make a version of the patient day table that includes the microbiome data after domination
tblpatientday_clinicalmicro_dominated <- tblpatientday_clinical_dominated %>%
    mutate(DayRelativeToNearestHCT = as.numeric(as.character(DayRelativeToNearestHCT))) %>%
    left_join(tblrel_genus_meta, by = c("PatientID", "DayRelativeToNearestHCT")) %>%
    mutate(DayRelativeToNearestHCT = as.factor(DayRelativeToNearestHCT)) %>%
    filter(Enterococcus_abund >= 0.1 & WBCtotal < 0.5 & Enterococcus_infection.y == F) %>%
    drop_na()
```

Let's check the effect of the most important classes of antibiotics and clinical covariates on *E. faecium* infection across the full clinical dataset:

```{r}
# Fit logistic regression model to full clinical dataset
clinical_model <- glm(
    Enterococcus_faecium_infection ~ WBCtotal + penicillins + carbapenems + cephalosporins + `glycopeptide antibiotics` + quinolones + DayRelativeToNearestHCT,
    data = tblpatientday_clinical,
    family = binomial(link = "logit")
)

# Summary of the model
summary(clinical_model)
```

Let's make sure that the effect of the antibiotics remains qualitatively equivalent when the reference day is the day before infection to remove cases where the antibiotics are administered reactively:

```{r}
# Data table with infection shifted
tblpatientday_clinical_shifted <- tblpatientday_clinical %>%
  arrange(PatientID, DayRelativeToNearestHCT) %>%
  group_by(PatientID) %>%
  mutate(Enterococcus_infection_shifted = lag(Enterococcus_infection, n = 1, default = 0)) %>%
  mutate(Enterococcus_infection_shifted = if_else(
    Enterococcus_infection_shifted == 1 & cumsum(Enterococcus_infection_shifted) == 1, 1, 0
  )) %>%
  ungroup()

# Fit logistic regression model to full clinical dataset
clinical_model_shifted <- glm(
    Enterococcus_infection_shifted ~ WBCtotal + penicillins + carbapenems + cephalosporins + `glycopeptide antibiotics` + quinolones + DayRelativeToNearestHCT,
    data = tblpatientday_clinical_shifted,
    family = binomial(link = "logit")
)

# Summary of the model
summary(clinical_model_shifted)
```

Now, we fit a similar model but including the abundance of *E. faecium*

```{r}
# Fit logistic regression model to full clinical dataset
microbiome_model <- glm(
    Enterococcus_infection.x ~ WBCtotal + penicillins + carbapenems + cephalosporins + DayRelativeToNearestHCT,
    data = tblpatientday_microbiome,
    family = binomial(link = "logit")
)

# Summary of the model
summary(microbiome_model)
```

And whether Enterococcus relative abundance is a predictor before and after domination

```{r}
# Fit logistic regression model
relabund_model <- glm(
    Enterococcus_infection.x ~ log10(Enterococcus_abund + 0.0001) + WBCtotal + DayRelativeToNearestHCT,
    data = tblpatientday_clinicalmicro_dominated,
    family = binomial(link = "logit")
)

# Summary of the model
summary(relabund_model)
```

Let's see

```{r}
# Patient-day table with micro data
tblpatientday_clinicalmicro <- tblpatientday_clinical %>%
  mutate(DayRelativeToNearestHCT = as.numeric(as.character(DayRelativeToNearestHCT))) %>%
  left_join(tblrel_genus_meta, by = c("PatientID", "DayRelativeToNearestHCT")) %>%
  left_join(domination_bsi_summary, by = "PatientID") %>%
  filter(Enterococcus_abund > 0.1) %>%
  drop_na()

  # Compute per-patient maximum Enterococcus abundance for dominated patients and plot densities + histogram
  patient_max_abund <- tblpatientday_clinicalmicro %>%
    filter(ever_dominated == TRUE) %>%
    group_by(PatientID, ever_infected) %>%
    summarize(max_abund = max(Enterococcus_abund, na.rm = TRUE), .groups = "drop") %>%
    drop_na(max_abund) %>%
    mutate(ever_infected = factor(ever_infected, levels = c(FALSE, TRUE),
                                  labels = c("Never infected", "Ever infected")))

  library(ggplot2)

  # Density plot: two densities (ever infected vs never infected) on same plot
  p_density <- ggplot(patient_max_abund, aes(x = max_abund, color = ever_infected, fill = ever_infected)) +
    geom_density(alpha = 0.35) +
    labs(
      x = "Maximum Enterococcus relative abundance (per patient)",
      y = "Density",
      color = "",
      fill = "",
      title = "Distribution of maximum Enterococcus abundance\namong ever-dominated patients, by infection status (density)"
    ) +
    theme_minimal()

  print(p_density)

  # Histogram plot: counts on y-axis (number of patients)
  medians <- patient_max_abund %>%
    group_by(ever_infected) %>%
    summarize(med = median(max_abund, na.rm = TRUE), .groups = "drop")

  p_hist <- ggplot(patient_max_abund, aes(x = max_abund, fill = ever_infected)) +
    geom_histogram(position = "identity", alpha = 0.85) +
    geom_vline(data = medians, aes(xintercept = med, color = ever_infected), linetype = "dashed", size = 1) +
    scale_fill_manual(
      values = c(
        "Ever infected" = "#0D7E2B", "Never infected" = "gray75"
      )
    ) +
    scale_color_manual(
      values = c(
        "Ever infected" = "#0D7E2B", "Never infected" = "gray75"
      )
    ) +
    labs(
      x = "Max. Enterococcus relative abundance (per patient)",
      y = "Number of patients",
      color = "",
      fill = "Infection status"
    ) +
    custom_theme +
    theme(
      legend.position = c(0.65, 0.85), # adjust as needed
      legend.background = element_rect(fill = "white", color = NA)
    )

  print(p_hist)

# Save the density and histogram plots as PDF
ggsave(filename = "document/plots/max_enterococcus_density.pdf",
  plot = p_hist,
  width = 4, height = 4)
```