---
title: "*Enterococcus* mouse experiments"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

## Pilot I - Colonization, expansion, and translocation

The experiment had three groups of mice:

  - Group A: received continuous oral ampicillin (CONCENTRATION) and *E. faecium* ATCC 700221 via oral gavage (CONCENTRATION)
  - Group B: received three doses of clindamycin (CONCENTRATION) and *E. faecium* ATCC 700221 via oral gavage (CONCENTRATION)
  - Group C: received three doses of clindamycin (CONCENTRATION) and *E. faecalis* V583 via oral gavage (CONCENTRATION)

Fecal samples, blood, and organ dilutions were plated on Enterococcosel agar (EA) and EV + vancomycin CONCENTRATION (EAV).

### Fecal CFU dynamics

Let's start by parsing the excel sheets with the fecal CFU counts:

```{r}
library(tidyverse)
library(readxl)

# List data sheet names
poop_sheets <- excel_sheets("data/lab_raw/mouse/pilot1/CFU_Exp1_Pilot.xlsx")[1:13]

# Read and combine all sheets, forcing "Notes" to character
poop_cfus <- lapply(poop_sheets, function(sheet) {
  df <- read_excel("data/lab_raw/mouse/pilot1/CFU_Exp1_Pilot.xlsx", sheet = sheet)
  df %>% mutate(Notes = as.character(Notes))
}) %>%
  bind_rows() %>%
  mutate(
    Date_parsed = as.Date(Date, format = "%m.%d.%y"),
    # Reference date: October 23, 2025, gavage date
    day = as.integer(Date_parsed - as.Date("2025-10-23")),
    cfu_per_g = (`Colony Number` * Dilution * 10) / `Weight of poop`,
    log10_cfu_per_g = log10(cfu_per_g + 1),
    Plate = str_replace(Plate, "EA`", "EA"),
    group = str_remove(Mouse_Name, "\\d+")) %>%
  rename(n_colonies = `Colony Number`,
        poop_weight = `Weight of poop`,
        species = `Bacteria Type`,
        mouse = Mouse_Name,
        date = Date,
        dilution = Dilution,
        media = Plate) %>%
  select(date, day, group, mouse, media, cfu_per_g,log10_cfu_per_g, dilution, n_colonies, poop_weight, species)

# Identify all mouse-day combinations with EA
ea_rows <- poop_cfus %>% filter(media == "EA")

# Find mouse-day combos missing EAV
missing_eav <- ea_rows %>%
  anti_join(
    poop_cfus %>% filter(media == "EAV") %>% select(mouse, day),
    by = c("mouse", "day")
  )

# Create filled EAV rows
filled_eav <- missing_eav %>%
  mutate(
    media = "EAV",
    n_colonies = 0,
    cfu_per_g = 0,
    log10_cfu_per_g = 0,
    dilution = 1
  )

# Bind filled rows to poop_cfus
poop_cfus <- bind_rows(poop_cfus, filled_eav) %>%
  arrange(day)
```

Now, we make a funciton that will summarize the poop CFUs for each group and plot the dynamics throughout the experiment:

```{r}
plot_poop_group <- function(poop_cfus, group_name) {
  # Filter for specified group and select entry with highest n_colonies per mouse per day and media
  poop_group <- poop_cfus %>%
    filter(group == group_name) %>%
    group_by(mouse, day, media) %>%
    slice_max(n_colonies, n = 1, with_ties = FALSE) %>%
    ungroup()
  
  # Calculate summary statistics for each day and media
  summary_group <- poop_group %>%
    group_by(day, media) %>%
    summarise(
      mean_log10 = mean(log10_cfu_per_g, na.rm = TRUE),
      q2.5 = quantile(log10_cfu_per_g, 0.025, na.rm = TRUE),
      q97.5 = quantile(log10_cfu_per_g, 0.975, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Create plot
  p <- ggplot(summary_group, aes(x = day, y = mean_log10, color = media, fill = media)) +
    geom_ribbon(aes(ymin = q2.5, ymax = q97.5), alpha = 0.2, color = NA) +
    geom_line() +
    geom_point(size = 2) +
    labs(
      x = "Day relative to gavage",
      y = expression("CFU/g"),
      color = "Media",
      fill = "Media"
    ) +
    scale_x_continuous(breaks = scales::breaks_width(2)) +
    scale_y_continuous(
      limits = c(0, 10),
      breaks = seq(0, 10, by = 2),
      labels = function(x) parse(text = paste0("10^", x))
    ) +
    #scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, by = 2)) +
    scale_color_manual(values = c("EA" = "#a08413", "EAV" = "#0D7E2B")) +
    scale_fill_manual(values = c("EA" = "#a08413", "EAV" = "#0D7E2B")) +
    #theme_minimal() +
    theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.759),
          panel.background = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(0.7, 0.9),
          text = element_text(color = "black"),
          axis.text = element_text(color = "black"))
  
  return(list(plot = p, poop_group = poop_group))
}
```

And the we generate the summaries and plots for each group:

```{r}
# Summaries
poop_cfus_A_sum <- plot_poop_group(poop_cfus, "A")$poop_group
poop_cfus_B_sum <- plot_poop_group(poop_cfus, "B")$poop_group
poop_cfus_C_sum <- plot_poop_group(poop_cfus, "C")$poop_group

# Plots
poop_cfus_A_plot <- plot_poop_group(poop_cfus, "A")$plot
poop_cfus_B_plot <- plot_poop_group(poop_cfus, "B")$plot
poop_cfus_C_plot <- plot_poop_group(poop_cfus, "C")$plot

# Export plots
ggsave("document/plots/poop_cfus_A_pilot1.pdf", poop_cfus_A_plot,
        width = 4.3, height = 3)
ggsave("document/plots/poop_cfus_B_pilot1.pdf", poop_cfus_B_plot,
        width = 4.3, height = 3)
ggsave("document/plots/poop_cfus_C_pilot1.pdf", poop_cfus_C_plot,
        width = 4.3, height = 3)
```

### Organ CFU results

Let's parse the counts for the organs and blood. Keep in mind that the liver, kidneys, lymph nodes, and mesenteric fat were often split before plating, so I took averages of the CFUs when processing the data: 

```{r}
# List data sheet names
organ_sheets <- excel_sheets("data/lab_raw/mouse/pilot1/CFU_Exp1_Pilot.xlsx")[14:17]

# Read and combine all sheets, forcing "Notes" and colony counts to character
organ_cfus_pre <- lapply(organ_sheets, function(sheet) {
  df <- read_excel("data/lab_raw/mouse/pilot1/CFU_Exp1_Pilot.xlsx", sheet = sheet)
  df %>% mutate(Notes = as.character(Notes), `Colony Number` = as.character(`Colony Number`))
}) %>%
  bind_rows() %>%
  mutate(
    `Colony Number` = as.numeric(str_replace(`Colony Number`, "excess", "1000")),
    cfu_per_unit = case_when(
                          Organ == "Blood" ~ `Colony Number` * Dilution * 10,
                          `Colony Number` == 0 & Dilution == 1 ~ 0,
                          TRUE ~ (`Colony Number` * Dilution * 10) / `Weight of poop`
                            ),
    Plate = str_replace(Plate, "EA`", "EA"),
    group = str_remove(Mouse_Name, "\\d+"))

# Get averages for organs processed in batches and calculate log10 CFUs    
organ_cfus_mean <- organ_cfus_pre %>%
      mutate(
        organ_category = case_when(
          str_detect(Organ, "Liver") ~ "mean_liver",
          str_detect(Organ, "Kidney") ~ "mean_kidney",
          Organ %in% c("MF 1", "Adipose 1", "Adipose", "MF 2") ~ "mean_mf",
          Organ %in% c("LN", "Lymph Node") ~ "mean_ln",
          TRUE ~ Organ
        )
      ) %>%
      group_by(Mouse_Name, Plate, organ_category) %>%
      summarise(cfu_per_unit = mean(cfu_per_unit, na.rm = TRUE), .groups = "drop") %>%
      mutate(
        log10_cfu_per_unit = log10(cfu_per_unit + 1),
        group = str_remove(Mouse_Name, "\\d+")
      ) %>%
      rename(
        mouse = Mouse_Name,
        media = Plate
      )

# Identify all mouse-media-organ_category combos with EA
ea_rows <- organ_cfus_mean %>% filter(media == "EA")

# Find combos missing EAV
missing_eav <- ea_rows %>%
  anti_join(
    organ_cfus_mean %>% filter(media == "EAV") %>% select(mouse, group, organ_category),
    by = c("mouse", "group", "organ_category")
  )

# Create filled EAV rows
filled_eav <- missing_eav %>%
  mutate(
    media = "EAV",
    cfu_per_unit = 0,
    log10_cfu_per_unit = 0
  )

# Bind filled rows to organ_cfus
organ_cfus <- bind_rows(organ_cfus_mean, filled_eav) %>%
  arrange(mouse, group, organ_category, media)
```

Now, we make a funciton to plot the CFUs for each organ in the two media:

```{r}
plot_organ_boxplot <- function(organ_cfus, organ_category, media) {
  fill_color <- if (media == "EA") "#a08413" else "#0D7E2B"
  
  organ_cfus %>%
    filter(organ_category == !!organ_category, media == !!media) %>%
    ggplot(aes(x = group, y = log10_cfu_per_unit, fill = group)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.6, fill = fill_color) +
      geom_jitter(width = 0.3, height = 0, size = 1.7) +
      labs(
        x = "Group",
        y = "CFU per unit"
      ) +
      scale_y_continuous(
        limits = c(0, 9),
        breaks = seq(0, 9, by = 1),
        labels = function(x) parse(text = paste0("10^", x))
      ) +
      theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.75),
            panel.background = element_blank(),
            legend.position = "none",
            text = element_text(color = "black"),
            axis.text = element_text(color = "black"))
}
```

And then plot and arrange all the organ CFUs:

```{r}
library(ggpubr)

# Blood plots
blood_ea_plot <- plot_organ_boxplot(organ_cfus, "Blood", "EA")
blood_eav_plot <- plot_organ_boxplot(organ_cfus, "Blood", "EAV")

# Organ plots
heart_ea_plot <- plot_organ_boxplot(organ_cfus, "Heart", "EA")
heart_eav_plot <- plot_organ_boxplot(organ_cfus, "Heart", "EAV")
liver_ea_plot <- plot_organ_boxplot(organ_cfus, "mean_liver", "EA")
liver_eav_plot <- plot_organ_boxplot(organ_cfus, "mean_liver", "EAV")
kidney_ea_plot <- plot_organ_boxplot(organ_cfus, "mean_kidney", "EA")
kidney_eav_plot <- plot_organ_boxplot(organ_cfus, "mean_kidney", "EAV")
spleen_ea_plot <- plot_organ_boxplot(organ_cfus, "Spleen", "EA")
spleen_eav_plot <- plot_organ_boxplot(organ_cfus, "Spleen", "EAV")
mf_ea_plot <- plot_organ_boxplot(organ_cfus, "mean_mf", "EA")
mf_eav_plot <- plot_organ_boxplot(organ_cfus, "mean_mf", "EAV")

# Small intestine plots
duo_ea_plot <- plot_organ_boxplot(organ_cfus, "Duodenum", "EA")
duo_eav_plot <- plot_organ_boxplot(organ_cfus, "Duodenum", "EAV")
jeju_ea_plot <- plot_organ_boxplot(organ_cfus, "Jejunum", "EA")
jeju_eav_plot <- plot_organ_boxplot(organ_cfus, "Jejunum", "EAV")
ile_ea_plot <- plot_organ_boxplot(organ_cfus, "Ileum", "EA")
ile_eav_plot <- plot_organ_boxplot(organ_cfus, "Ileum", "EAV")

# Large intestine plots
cecum_ea_plot <- plot_organ_boxplot(organ_cfus, "Cecum", "EA")
cecum_eav_plot <- plot_organ_boxplot(organ_cfus, "Cecum", "EAV")
cecumcon_ea_plot <- plot_organ_boxplot(organ_cfus, "Cecum Content", "EA")
cecumcon_eav_plot <- plot_organ_boxplot(organ_cfus, "Cecum Content", "EAV")
colon_ea_plot <- plot_organ_boxplot(organ_cfus, "Colon", "EA")
colon_eav_plot <- plot_organ_boxplot(organ_cfus, "Colon", "EAV")
coloncon_ea_plot <- plot_organ_boxplot(organ_cfus, "Colon Content", "EA")
coloncon_eav_plot <- plot_organ_boxplot(organ_cfus, "Colon Content", "EAV")

# Arrange plots for slides
si_plots <- ggarrange(
                      duo_ea_plot, jeju_ea_plot, ile_ea_plot,
                      duo_eav_plot, jeju_eav_plot, ile_eav_plot,
                      nrow = 2, ncol = 3
                      )
li_plots <- ggarrange(
                      cecum_ea_plot, colon_ea_plot,
                      cecum_eav_plot, colon_eav_plot,
                      nrow = 2, ncol = 2
                      )
organ_plots <- ggarrange(
                      mf_ea_plot, blood_ea_plot, liver_ea_plot, heart_ea_plot, kidney_ea_plot, spleen_ea_plot,
                      mf_eav_plot, blood_eav_plot, liver_eav_plot, heart_eav_plot, kidney_eav_plot, spleen_eav_plot,
                      nrow = 2, ncol = 6
                      )

# Save plots
ggsave("document/plots/organ_cfus_pilot1.pdf", organ_plots, width = 10, height = 5)
ggsave("document/plots/si_cfus_pilot1.pdf", si_plots, width = 5, height = 5)
ggsave("document/plots/li_cfus_pilot1.pdf", li_plots, width = 3.34, height = 5)
```